\section{Changes so far}

\tom{this section is for internal circulation and will be omitted in the student version}

General changes include:
\begin{itemize}
\item \textbf{Formatting:} Used environments for theorems, definitions, examples, exercise, and solutions. Each visually different so that it works with black/white print.
\item \textbf{Wording:} Removed words that cause ambiguity, such as \emph{try} or \emph{guess}. Removed words such as \emph{clearly} or \emph{easily}.
\item \textbf{Details:} Included addition definitions/theorems to separate, clarify ideas, and improve mathematical rigour.
\item \textbf{Updating tex:} Mostly behind the scenes stuff. Convert tex to more use modern environments. Try to use less manual spacing
\item \textbf{Pictures:} Converted graphs to tikz (where possible)
\item \textbf{Notation:} Changed $\sin^{-1}$ to $\arcsin$. Consistency with bracketing
\item \textbf{Hyperlink elements:} when referring to previously mentioned examples and general navigation
\item \textbf{QR codes:} for increased accessibility to interactive elements (mostly desmos).
\item \textbf{Typesetting style:} Exercise solutions formatting to resemble written solutions
\item \textbf{Footnotes:} Added in places for relevant additional notes or caveats
\item \textbf{Tutorial:} Questions added at the end of relevant sections. Answers at the end of document. Questions/Answers hyperlinked.
\end{itemize}

\subsection{Chapter 1: Applications of Differentiation}
\begin{itemize}
\item Added point-slope formula.
\item The \question symbol to means solutions left for lectures or as an exercise.
\item Ordering change: Product rule, quotient rule, chain rule (consistent with F17XA)
\item Ordering change: Linear approximation before Maclaurin and Taylor
\item Introduced notation $L(x)$ for linear approximation
\item Differentiate between Polynomial/series
\item Ordering change: polynomial approximation before series
\item \textbf{Do we need 2nd derivative test?}
\item Optimisation: Steps introduced
\item Defined differential equations and provided examples
\end{itemize}

\subsection{Chapter 2: Techniques of Integration}
\begin{itemize}
\item $\int \frac{1}{x} dx = \ln(x) + C$ or $\int \frac{1}{x} dx = \ln|x| + C$ ?
\item Reordering: Partial fractions, substitution, then by parts.
\end{itemize}


\subsection{Chapter 3: Applications of Integration}
\begin{itemize}
\item Initial value problem: included $\pm \sqrt{\cdot}$ solutions.
\end{itemize}


\subsection{Chapter 4: Complex number}
\begin{itemize}
\item Changed ordering of conjugates before division
\item Changed diagrams to tikz
\item Inequalities in complex numbers (\autoref{ex:compareas})
\item Omitted \emph{cis} notation.
\item Diagrams associated with \autoref{eqn:comptrans} converted to tikz. That Aerofoil picture... ARGH!
\end{itemize}

\subsection{Chapter 4: Matrices}
\begin{itemize}
\item Reordering: Matrix definitions, equality, transpose, types of matrices. Then Arithemtic
\item Introduced $s(a_{ij})$ as the sign of element $a_{ij}$.
\end{itemize}
